---
title: Full title of the paper (Capitalized)
author:
  - name: Dominik Leutnant
    affil: 1,2,\ddagger,*
    orcid: 0000-0003-3293-2315
  - name: John Doe
    affil: 2, \dagger, \ddagger
affiliation:
  - num: 1
    address: |
      Muenster University of Applied Sciences - 
      Institute for Infrastructure, Water, Resources, Environment
      Correnstr. 25, 48149 Muenster, Germany
    email: leutnant@fh-muenster.de
  - num: 2
    address: |
      Your department
      Street, City, Country
    email: mail@mail.com
# author citation list in chicago format
authorcitation: |
  Leutnant, D.; Doe, J.
# firstnote to eighthnote
firstnote: |
  Current address: Updated affiliation
secondnote: |
  These authors contributed equally to this work.
correspondence: |
  leutnant@fh-muenster.de; Tel.: +XX-000-00-0000.
# document options
journal: notspecified
type: article
status: submit
# front matter
simplesummary: |
  A Simple summary goes here.
abstract: |
  A single paragraph of about 200 words maximum. For research articles, 
  abstracts should give a pertinent overview of the work. We strongly encourage
  authors to use the following style of structured abstracts, but without 
  headings: 1) Background: Place the question addressed in a broad context and
  highlight the purpose of the study; 2) Methods: Describe briefly the main
  methods or treatments applied; 3) Results: Summarize the article's main 
  findings; and 4) Conclusion: Indicate the main conclusions or interpretations. 
  The abstract should be an objective representation of the article, it must not 
  contain results which are not presented and substantiated in the main text and 
  should not exaggerate the main conclusions.
# back matter
keywords: |
  keyword 1; keyword 2; keyword 3 (list three to ten pertinent keywords specific 
  to the article, yet reasonably common within the subject discipline.).
acknowledgement: |
  All sources of funding of the study should be disclosed. Please clearly 
  indicate grants that you have received in support of your research work. 
  Clearly state if you received funds for covering the costs to publish in open 
  access.
authorcontributions: |
  For research articles with several authors, a short paragraph specifying their 
  individual contributions must be provided. The following statements should be 
  used ``X.X. and Y.Y. conceive and designed the experiments; X.X. performed the 
  experiments; X.X. and Y.Y. analyzed the data; W.W. contributed 
  reagents/materials/analysis tools; Y.Y. wrote the paper.'' Authorship must be
  limited to those who have contributed substantially to the work reported.
funding: |
  Please add: ``This research received no external funding'' or ``This research 
  was funded by NAME OF FUNDER grant number XXX.'' and  and ``The APC was funded 
  by XXX''. Check carefully that the details given are accurate and use the 
  standard spelling of funding agency names at 
  \url{https://search.crossref.org/funding}, any errors may affect your future 
  funding.
institutionalreview: |
  In this section, you should add the Institutional Review Board Statement and 
  approval number, if relevant to your study. You might choose to exclude 
  this statement if the study did not require ethical approval. Please note 
  that the Editorial Office might ask you for further information. Please add 
  “The study was conducted in accordance with the Declaration of Helsinki, 
  and approved by the Institutional Review Board (or Ethics Committee) of 
  NAME OF INSTITUTE (protocol code XXX and date of approval).” for studies 
  involving humans. OR “The animal study protocol was approved by the 
  Institutional Review Board (or Ethics Committee) of NAME OF INSTITUTE 
  (protocol code XXX and date of approval).” for studies involving animals. 
  OR “Ethical review and approval were waived for this study due to REASON 
  (please provide a detailed justification).” OR “Not applicable” for
   studies not involving humans or animals.
informedconsent: |
  Any research article describing a study involving humans should contain this 
  statement. Please add ``Informed consent was obtained from all subjects 
  involved in the study.'' OR ``Patient consent was waived due to REASON 
  (please provide a detailed justification).'' OR ``Not applicable'' for 
  studies not involving humans. You might also choose to exclude this statement 
  if the study did not involve humans.
  
  Written informed consent for publication must be obtained from participating 
  patients who can be identified (including by the patients themselves). Please 
  state ``Written informed consent has been obtained from the patient(s) to 
  publish this paper'' if applicable.
dataavailability: |
  We encourage all authors of articles published in MDPI journals to share 
  their research data. In this section, please provide details regarding where 
  data supporting reported results can be found, including links to publicly 
  archived datasets analyzed or generated during the study. Where no new data 
  were created, or where data is unavailable due to privacy or ethical 
  re-strictions, a statement is still required. Suggested Data Availability 
  Statements are available in section “MDPI Research Data Policies” at 
  \url{https://www.mdpi.com/ethics}.
conflictsofinterest: |
  Declare conflicts of interest or state 'The authors declare no conflict of 
  interest.' Authors must identify and declare any personal circumstances or
  interest that may be perceived as inappropriately influencing the
  representation or interpretation of reported research results. Any role of the
  funding sponsors in the design of the study; in the collection, analyses or 
  interpretation of data in the writing of the manuscript, or in the decision to 
  publish the results must be declared in this section. If there is no role, 
  please state 'The founding sponsors had no role in the design of the study; 
  in the collection, analyses, or interpretation of data; in the writing of the 
  manuscript, an in the decision to publish the results'.
sampleavailability: |
  Samples of the compounds ...... are available from the authors.
supplementary: |
 The following supporting information can be downloaded at:  
 \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title.
abbreviations:
  - short: MDPI
    long: Multidisciplinary Digital Publishing Institute
  - short: DOAJ
    long: Directory of open access journals
  - short: TLA
    long: Three letter acronym
  - short: LD 
    long: linear dichroism
bibliography: mybibfile.bib
appendix: appendix.tex
endnotes: false

output:
  rticles::mdpi_article:
    extra_dependencies: longtable
---

```{r mdpitype, echo = FALSE}
type <- c("abstract, addendum, article, book, bookreview, briefreport, 
casereport, comment, commentary, communication, conferenceproceedings, 
correction, conferencereport, entry, expressionofconcern, extendedabstract, 
datadescriptor, editorial, essay, erratum, hypothesis, interestingimage, 
obituary, opinion, projectreport, reply, retraction, review, perspective, 
protocol, shortnote, studyprotocol, systematicreview, supfile, 
technicalnote, viewpoint, guidelines, registeredreport, tutorial")
knitr::kable(rticles::string_to_table(type, 3), 
             col.names = NULL, format = "latex", booktab = TRUE, 
             caption = "MDPI article types.")


```

```{r, setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

```

# Introduction

## POMDP

## Simulation details

# Results

## Performance of POMDPs vs ADT heuristic

```{r}
#set colors
adt_col <- "green"
long_col <- "red"
medium_col <- "skyblue"
short_col = "violet"

strat_colors <- c("ADT" = adt_col, "Long" = long_col, "Medium" = medium_col, "Short" = short_col)
contrast_colors <- c("long_vs_adt" = long_col, "medium_vs_adt" = medium_col, "short_vs_adt" = short_col)


#print densities
source("sim_lengths_densities_full.R")

plot + 
  #set color scheme so that long _vs_adt is green
  scale_fill_manual(values = strat_colors) +
  scale_color_manual(values = strat_colors) 



```

Four different simulation runs each consisting of 100 runs, of maximally 200 timesteps for different rates of tumor growth (respectively 1. .3, .5, and .7 probability of increasing the tumor state) . On each run the the outcomes on each of the potential 200 random variables are generated and each strategy is therefore tested on a the same enviroment. On the lowest tumor growth rate almost all the runs reach maximal length.

Contrasts against ADT-performance for growth rate .3, .5 and .7 are plotted. .1 is omitted since it appears that maximal perforamnce was achieved for each policy horizon of the POMPD strategies.

```{r}
source("contrasted_sim_lengths_.R")

plot +
  scale_fill_manual(values = contrast_colors) +
  scale_color_manual(values = contrast_colors) 
  
```

Two interpretations emerge from the contrasted simulation lengths. As the the growth rate increases, the POMDP seems to run for longer than the specific instance of the range bounded approach. It also seems that longer policy horizon seems beneficial. When the tumor state had .7 probability of increasing each timestep, the longest POMPD with the longest policy horizon performed as follows.

```{r}
source("contrasted_sim_length_max_horizon.R")

plot +
  scale_fill_manual(values = contrast_colors) +
  scale_color_manual(values = contrast_colors) 
```

At more aggressive cancer rates, the simulation consistently to outperform the range bounded approach. The absoulute increase in timesteps decreases however.

## Capabilities

### Learning underlying hidden state

The POMDP structure is capable of inferring an underlying hidden state: the resistance state. Even though this not directly observable, this can be inferred to how the tumor state responds to treatment. Applying treatment for longer time without any beneficial effect would suggest that the resistance level is high, while immediately observing that the tumor level decreases would suggest that the resistance level is low. At each timestep the POMDPs perform infer the most likely state of every state factor, given their current observation and prior beliefs. Through custom changes to the 'get_expected_states' function in PYMDP that allowed the models to consider how one hidden state factor (resistance factor) would influence another (the tumor factor).

```{r}
source("few_time_steps_res_beliefs.R")

plot

```

The above plot shows the strength of belieffs in t probabilities, and the red dot show the actual the resistance levels. A time progresses the resistance level increases, and the model adjusts its beliefs. While the tumor level doesn't increase durings this time_period, this would also signal the model that the resistance level is low. This is the case since the enviroment always has change of increasing the tumor state, but such an increase would be negated by succesful round of treatment.

```{r}
source("sim_run_example.R")

plot

```

This updating corresponds to a period in the given simulation run the where the POMDP is choosing to treat, and it there

```{r}
source("resistance_beliefs_plot.R")
plot
```

### Beliefs, current/future and unceartainty is accessible

### Actions Can be balanced using FEP

### FEP components are accesible

# Discussion

## Does it work for other sorts of medical planning

## Would longer policy search be better

## Only one range bound

## Future Research

### Continuous state space or binning illness

### Learning the transition parameters

### searching policy space better

### other fep models
